Upwelling variability across the Benguela upwelling systems

### 2019-09-23 Write_up/Paper1_writeup.Rmd
Comments throughout the text thus far. My comments begin with RWS, for ease of searching.

### Introduction

Upwelling is a physical oceanographic process in which cold, nutrient rich waters are pushed to the surface as a result of the motion caused by alongshore equatorial winds. These upwelling events are known to have a large influence on the abundance, diversity, and distribution of marine organisms, and moreover, critically affect marine ecosystem processes (Add in references). Globally, upwelling occurs throughout the world’s oceans, however, four major eastern boundary upwelling systems (EBUS) are recognised, each significantly impacting coastal systems. These respective systems are characterized as vast regions of coastal ocean occurring along the western shores of continents bordering the Pacific and Atlantic Oceans [@bakun; @pauly]. The oceanography of the inshore waters off the west coast of South Africa is dominated by an upwelling system which, along with the Peru, California and Humboldt currents as Eastern boundary upwelling regions [@andrews]. Eastern boundary upwelling regions only cover 1% of the world’s ocean surface but are among the most productive regions of the world. As such, there has been a constant interest in upwelling systems. Major upwelling programs have already been carried through off of Peru (e.g. WALSH. KELLEY, DUGDALE and FROST, 1971) and North-West Africa (e.g. BARBER, 1977).

Over the past three decades, climate change as a result of anthropogenic activities has established global concern and is shown to have negatively affected marine and terrestrial ecosystems. The consequences of these changes are also shown to have direct effects on the behaviour and functioning of marine organisms, thus ultimately affecting species trophic levels and integration [@doney]. According to the Bakun hypothesis, an increase in greenhouse gases will result in an increase in day-time warming and night-time cooling and ultimately cause an increase in temperature gradients which will then form stronger pressure gradients. Pressure gradients are known to affect wind patterns which will subsequently affect the intensity and duration of upwelling events [@hsieh; @mote; @lima].

Several studies have provided significant evidence suggesting that there has been an increase in ocean surface temperatures over the last few decades (Add references). These temperatures have increased as a result of anthropogenically driven climate change [@gruber]. Warming is shown to slowly move down into the water column thus increasing thermal stratification. This thermal stratification restricts the depth at which water is upwelled and therfore limits the amount of nutrients brought up to the surface with a decrease in oxygen and pH levels [@chhak; @jacox2011; @jocox2016]. Climate change impacting oceanic processes has major ecological consequences (Bradley et al. 1999; Walther et al. 2002). This is demonstrated in several regions increases in sea surface temperatures have been proven to greatly influence marine species survival, reproduction and abundance. Climate change has also been shown to largely influence coastal species such as corals, muscles and oysters. Given the importance of primary production, it was determined that the Benguela upwelling system, the eastern boundary upwelling system in South Africa, should be the focus of this paper. Furthermore, it should be understood that increases in sea surface temperatures bear considerable economic impacts as well (Murawski 1993) therefore understanding the impact and cause of climate change has merit not only in mitigation efforts in limiting environmental degradation and maintaining ecosystem functioning, but in its negative affects to local and broader economies as well as human populations dependent on these coastal regions.

The 3000km South African coastline, is bordered by the Agulhas current along the east coast and the EBUS Benguela current, along the west coast. These coastal regions display noticeable differences in seawater temperatures in comparison to each other, primarily due to the influences of the neighbouring ocean currents (Bolton et al. 2004; Mead et al. 2013; Schlegel and Smit 2016). The Benguela Current is driven by the anticyclone high pressure systems and eastward moving cyclones which determine the boundaries of the Benguela upwelling region (Shannon 2006; Hutchings et al. 2009). The atmospheric temperatures along the coastline are largely influenced by the Benguela Current [@van], however, upwelling systems may be forced by various atmospheric conditions [@Lutjeharms; @roberts; @hutchings]. Regionally, the Benguela Current assists in transporting cold water northwards from the Southern Ocean to the coast (Lüning 1990; Lutjeharms, Cooper, and Roberts 2000; Hutchings et al. 2009; Schlegel et al. 2017). Changes in the intensity and duration of upwelling events have proven to cause large changes in ecosystem functioning [@wang].

The Benguela system is bounded by the equatorward and poleward ends by warm water regimes. These currents are known to be forced by local wind stress fields off southwest Africa [@nelson; @fennel]. This particular EBUS stretches along the south-western coast of Africa, from southern Angola to Cape Agulhas [@cole], with water temperatures coldest and most persistent off the Namibian coast at 25°S and weakest at 17°S [@Lutjeharms]. The upwelling period in the Benguela north of 25°C persists throughout the entire year but peaks during August. The Benguela upwelling region is split into northern and southern systems by a zone of intense perennial upwelling activity in the Luderitz region within the Namibian region [@shannon; @cole]. Several studies have shown that extreme warming and cooling periods occurring in the northern and southern Benguela systems [@stander; @walker; @shannon1984; @boyd; @brainard]. Furthermore, Walter (1937) found that warm periods of global ocean temperatures may have resulted as a lack of upwelling or poor upwelling events. During the periods of warming, the equatorward wind stress over the northern region of the Benguela system was stronger than usual yet upwelling was suppressed by the southward movement of warm saline waters from the Angolan region. In the southern Benguela region, continental shelf bathymetry and strong upwelling favourable winds are known to result in more intense upwelling events. Upwelling occurring along the south coastline of the Southern Africa region are usually generated by local winds. Event-scale upwelling has been demonstrated to have substantial impacts on fisheries [@roy; @van].

The Cape peninsula is influenced by south-east trade winds during summer and westerlies during winter, thus a pronounced seasonal cycle is evident within patterns of wind movement across the coastline. During spring, summer, and autumn seasons, the anticyclonic high-pressure cell periodically ridges eastwards south of the continent. During this process, south easterly winds predominate, with peak frequencies occurring during spring and summer months resulting in more upwelling events during summer and winter months. During the winter seasons, low pressure cells are driven from the west to the east as a result of the atmospheric circumpolar waves and so north-westerly winds dominate during winter. South easterly winds dominate during summer but are still present for short durations during winter [@andrews]. Within a range of kilometres off the shore, wind conditions vary due to the effects of mountain funnelling within the region.

The dynamics of upwelling within the nearshore (<400m from the coastline) remains poorly understood. Upwelling metrics are often calculated far offshore because measurements are typically taken from large ships that cannot move close to the shore [@andrews]. Cold temperatures are important in describing species population distribution limits and are proven to set limits in the north or southwards higher latitudes (Firth et al., 2011). There has not yet been any understanding of how the upwelling process varies at local versus broader scales. In this study, it was hypothesize that upwelling is more dominant within coastal regions instead of broader scale. Very few studies have addressed the changes in the intensity or duration of cold temperature spells or cold events at varying distances from the coast within the Benguela region.

Strong temperature gradients existing between offshore and inshore oceanic waters may result in different upwelling event readings. Upwelling relatively close to the shore has been found to depend on the strength and direction of the wind and tends to vary based on seasonal wind variation [@andrews]. Inshore temperatures tend to be lower during summer months due to upwelling, but there is an increase in offshore temperatures in the later summer months as a result of warming of the South East Atlantic Surface waters[@andrews,@Shannon]. Due to the temperature gradients that exist inshore and offshore, this study focused on observing upwelling patterns from the coastal region to 50km from the coastline.

The central assumption of this project was that remotely-sensed SST data and *in situ* seawater temperature data may be compared. Remotely sensed SST data and *in situ* seawater temperature data were obtained at different resolutions and so slight changes may be found within the different datasets. The primary aim of the first study was to make use of various remotely sensed SST and in-situ collected coastal seawater temperature data. The most appropriate dataset for the measurement of the nearshore temperatures within South Africa was the South African Coastal Temperature Network (SACTN) dataset [@smit]. It was useful to examine whether or not the same upwelling patterns exist at different distances from the coastline and if these patterns that exist are evident within the various datasets. This investigation examined recent trends and variation of upwelling events at different distances from the coastline.

@@

## Site selection

On a broad scale, the South African coastline exhibits a large variation in seawater temperature and is divided into four bioregions, each with contrasting temperatures [@smit, @mead]. The western region of the coastline is dominated by the Benguela Current forming an Eastern Boundry Upwelling System (EBUS) [@hutchings], which provides a natural laboratory for this study. Annual mean coastal seawater temperature within this region represents a coastal seawater temperature range of 12.3 ± 1.2°C at the Western limit near the Namibian border. Seasonal upwelling is controlled by south-easterly trade winds and as a result reflects distinct temperature variations representing much lower temperatures within the upwelling cells over a fairly narrow continental shelf found from the Cape Peninsula to Cape Columbine. This region represents monthly mean temperatures at generally intermediate at cold and warm temperatures [@luning].  In order to accurately examine upwelling patterns at various distances from the coast, several sites were selected providing a more accurate representation. The appropriate sites along the west coast of South Africa (i.e. sites influenced by the Benguela current) were selected. Since temperature data were not evenly recorded for each of the sites, the entire dataset was narrowed to only those sites with a time series of longer than 30 years. This yielded a total of five sites within the Benguela region (Fig 1).

@@

## Temperature data sources

The South African Coastal Temperature Network (SACTN) data served as the *in situ* temperature dataset used in this study. This dataset consists of coastal seawater temperatures obtained from 129 sites along the South African coastline. The SACTN has collected SST data along the coastline of South Africa from as early as 1972 till present date, with contributions made from several organisations and governmental departments. As a result, temperatures within these datasets were obtained by using different methods and instruments, such as hand held thermometers or Underwater temperature reorders( UTRs). The precision of the instruments used vary accordingly, with hand-held thermometers recording temperatures at a precision of 0.5°C as opposed to the underwater temperature recorders which have an instrument precision of 0.001°C. The duration and extent of recordings per sites were uneven with the longest time series in the dataset being that of Gordons Bay which started on 13 September 1972 and concluded on 26 January 2017.  During the 1970s, a total of 11 time series began recording. A further 53 entries were added during the 1980s, 34 during the 1990s, and 18 entries were added during the 2000s. Recordings are still ongoing at many of these sites.

This study also made use of a range of remotely sensed temperature datasets. Satellites provide a system of large spatial and temporal data retrieval. Remotely sensed SST products are produced by a number of organisations across the globe. The AVHRR-only Optimally-Interpolated Sea Surface Temperature (OISST) was used to determine SST within the study region. The AVHRR datasets have been providing global SSTs for more than four decades (Reynolds and Smith, 1994; Pearce et al 2006). OISST is a global 1/4° gridded, daily SST product that assimilates both remotely sensed and in situ sources of data to create a level-4 gap free product (Banzon et al., 2016). The Multi-scale Ultra-high Resolution (MUR) Sea Surface Temperature Analysis, the second dataset, was produced using satellite instruments with datasets spanning 1 June 2002 to present times (PO.DAAC, https://podaac.jpl.nasa.gov/). MUR provides SST data at a spatial resolution of 0.01° in longitude-latitude coordinates and are currently among the highest resolution SST datasets available. The CMC dataset constitutes the third dataset and is a version 3.0 Group for High Resolution Sea Surface Temperature (GHRSST) Level 4 dataset with a 10km resolution constructed by the Canadian Meteorological Center (CMC). The CMC dataset combines infrared satellite SST at numerous points in the time series from the AVHRR, the European Meteorological Operational-A (METOP-A) and Operational-B (METOP-B) platforms, and microwave SST data from the Advanced Microwave Scanning Radiometer 2 in conjunction with in situ observations of SST from ships and buoys from the ICOADS program.

An advantage of using of *in situ* data over satellite data is that they provide a more accurate representation of the thermal properties closer to the coast where satellite temperatures do not capture temperature properties well and can therefore more accurately explain upwelling within the coastal inshore environment. Smit et al (2013) have shown that satellite data collected along the South African coastline represent a warm bias reaching as high as 6°C over *in situ* temperatures within the nearshore environment.

In order to create a time series of each of the remotely sensed SST data and also to compare whether the same upwelling patterns exist on the broad-scale against the *in situ* time series, shore-normal transects were extended from the *in situ* sites along the west coast of South Africa. Temperature values from the various satellite datasets were extracted at multiple distances from the coast. These distances were 15 x 15 km, 25 X 25 km and 50 X 50 km pixels along the transect (Fig. 1). The daily sea surface temperatures obtained from the various remotely sensed datasets within each of the transects were initiated in order to produce offshore, broad-scale time series that could accurately be used as a dataset to be compared against the *in situ* time series.

## Wind data

Wind action was an important variable in this study as it was hypothesised that wind direction and wind speed may have a direct influence on upwelling, consequently, wind action was investigated for its impact on upwelling at specific sites along the west coast of South Africa. Wind data was obtained from the South African Weather Service (SAWS), and were provided at three-hour resolutions. Specific wind characteristics were measured, namely; wind direction (dirw) and wind speed (spw). Since the wind data were modelled at three-hour resolutions, they were converted into daily data points in order to compare them with the temperature data. The circular() function in R software was then used to create circular objects around the wave data in order to calculate the daily wind parameters. ERA5 hourly data on pressure levels was used as a second wind source. This dataset comprised of u and v variables from 1973 to present.

With temperature and wind values now corresponding to their respective sites and dates, the hypothesis testing upwelling patterns at different distances from the coastline could now be tested. To do this, it was important to identify upwelling.

# Defining and determining upwelling

In order to determine the change in upwelling signals at different distances from the coastline and to test which dataset showed clear, visible representation of these events, it was first necessary to define when upwelling was occurring. The next aim was to determine a threshold value to identify when upwelling was occurring within this region. Given that upwelling is primarily caused by alongshore, equator ward winds, it was necessary to consider both temperature and wind values. To determine the upwelling condition, an upwelling index was calculated using the formula presented in Fielding and Davis (1989). The upwelling index provided an indication of the intensity of the upwelling process taking place using the following equation:

@@

In this expression, μ represents the wind speed (m/s) and θ represents the wind direction which is measured in degrees. The coastal angle is the angle of each of the sites perpendicular to the coastline. The index relies heavily on wind speed and direction data in order to determine the intensity of upwelling likely to occur. Thereafter, the 'exceedance' function within the heatwaveR package was used to determine the consecutive number of days at or above which the upwelling index value was ment to be. Here, it was specified that the minimum duration for upwelling was 1 day with a threshold value obtained from the upwelling index as 1. In order to obtain a more approved threshold value, a percentile value was determined. To determine changes within the upwelling values, the 'detect event' function within the heatwaveR package was used. Thereafter, to determine the upwelling event metrics ( frequency and intensity) the detect event function was used as a first filter with a threshold of 25th percentile. Values obtained from the upwelling index were used as a second threshold. An R script (https://github.com/AmierohAbrahams/Upwelling_MCS) was used to calculate the individual upwelling events for both the *in situ* and remotely sensed SST data. Because upwelling events were calculated relative to percentile exceedances, rather than an absolute definition such as periods with temperatures above a fixed temperature threshold, upwelling events occurred any time of the year with upwelling shown to be more dominant during spring and summer months. For the use of the heatwaveR package it was best to use a time series of 30years.

>>>>

## Introduction

Upwelling is a physical oceanographic process in which cold, nutrient rich waters are pushed to the surface as a result of the motion caused by alongshore equatorial winds. These upwelling events are known to have a large influence on the abundance, diversity, and distribution of marine organisms, and moreover, critically affect marine ecosystem processes (Add in references). Globally, upwelling occurs throughout the world’s oceans, however, four major eastern boundary upwelling systems (EBUS) are recognised, each significantly impacting coastal systems (RWS: This is a run-on sentence). These respective systems are characterized as vast regions of coastal ocean occurring along the western shores of continents bordering the Pacific and Atlantic Oceans [@bakun; @pauly]. The inshore waters off the west coast of South Africa are dominated by an upwelling system which, along with the Peru, California and Humboldt currents as Eastern boundary upwelling regions [@andrews] (RWS: This could be made more clear. What is the main point you want to communicate in this sentence?). Eastern boundary upwelling regions only cover 1% of the world’s ocean surface but are among the most productive regions of the world. As such, there has been a constant interest in upwelling systems. Major upwelling programs (RWS: This is too vague. What are these programs?) have already been carried out in Peru (e.g. WALSH. KELLEY, DUGDALE and FROST, 1971) and North-West Africa (e.g. BARBER, 1977) (RWS: Thes references need to be made in a markdown format (e.g. [@walsh])).

Over the past three decades, anthropogenic climate change has established global concern as it has been shown to have negatively affected marine and terrestrial ecosystems (RWS: citation needed). The consequences of these changes are also shown to have had direct effects on the behaviour and functioning of marine organisms, thus ultimately affecting species trophic levels and integration [@doney]. According to the Bakun hypothesis, an increase in greenhouse gases will result in an increase in day-time warming and night-time cooling and ultimately cause an increase in temperature gradients which will then form stronger pressure gradients [@bakun]. Pressure gradients are known to affect wind patterns which will subsequently affect the intensity and duration of upwelling events [@hsieh; @mote; @lima].

Several studies have provided significant evidence suggesting that there has been an increase in ocean surface temperatures over the last few decades (Add references; RWS: You can cite the IPCC for this). These temperatures have increased as a result of anthropogenically driven climate change [@gruber]. Warming is shown to slowly move down into the water column thus increasing thermal stratification (RWS: citation needed). This thermal stratification restricts the depth at which water is upwelled and therfore limits the amount of nutrients brought up to the surface with a decrease in oxygen and pH levels [@chhak; @jacox2011; @jocox2016]. Climate change impacting oceanic processes has major ecological consequences (Bradley et al. 1999; Walther et al. 2002 (RWS: Make these markdown citation style)). This is demonstrated in several regions increases in sea surface temperatures have been proven to greatly influence marine species survival, reproduction and abundance. Climate change has also been shown to largely influence coastal species such as corals, muscles and oysters (RWS: Citations). Given the importance of primary production, it was determined that the Benguela upwelling system, the eastern boundary upwelling system in South Africa, should be the focus of this paper. Furthermore, it should be understood that increases in sea surface temperatures bear considerable economic impacts as well (Murawski 1993 (RWS: Markdown citation)) therefore understanding the impact and cause of climate change has merit not only in mitigation efforts in limiting environmental degradation and maintaining ecosystem functioning, but in its negative affects to local and broader economies as well as human populations dependent on these coastal regions. (RWS: There appear to be two or three separate main points being made in this paragraph. I would recomend you split it up into a couple of paragraphs and think about what the main message of each should be.)

The 3000km South African coastline is bordered by the Agulhas current along the east coast and the Benguela current (an EBUS) along the west coast. These coastal regions display noticeable differences in seawater temperatures primarily due to the influences of the neighbouring ocean currents (Bolton et al. 2004; Mead et al. 2013; Schlegel and Smit 2016 (RWS: Markdown citations. Fix this throughout. I will stop making this comment here.)). The Benguela Current is driven by the anticyclone high pressure systems and eastward moving cyclones which determine the boundaries of the Benguela upwelling region (Shannon 2006; Hutchings et al. 2009). The atmospheric temperatures along the coastline are largely influenced by the Benguela Current [@van], however, upwelling systems may be forced by various atmospheric conditions [@Lutjeharms; @roberts; @hutchings]. Regionally, the Benguela Current assists in transporting cold water northwards from the Southern Ocean to the coast (Lüning 1990; Lutjeharms, Cooper, and Roberts 2000; Hutchings et al. 2009; Schlegel et al. 2017). Changes in the intensity and duration of upwelling events have proven to cause large changes in ecosystem functioning [@wang].

The Benguela system is bounded by the equatorward and poleward ends by warm water regimes. These currents are known to be forced by local wind stress fields off southwest Africa [@nelson; @fennel]. This particular EBUS stretches along the south-western coast of Africa, from southern Angola to Cape Agulhas [@cole], with water temperatures coldest and most persistent off the Namibian coast at 25°S and weakest at 17°S [@Lutjeharms]. The upwelling period in the Benguela north of 25°C persists throughout the entire year but peaks during August. The Benguela upwelling region is split into northern and southern systems by a zone of intense perennial upwelling activity in the Luderitz region in Namibia [@shannon; @cole]. Several studies have shown that extreme warming and cooling periods occur in the northern and southern Benguela systems [@stander; @walker; @shannon1984; @boyd; @brainard]. Furthermore, Walter (1937) found that warm periods of global ocean temperatures may have resulted as a lack of upwelling or poor upwelling events. During the periods of warming, the equatorward wind stress over the northern region of the Benguela system was stronger than usual yet upwelling was suppressed by the southward movement of warm saline waters from the Angolan region. In the southern Benguela region, continental shelf bathymetry and strong upwelling favourable winds are known to result in more intense upwelling events. Upwelling occurring along the southern coastline of southern Africa are usually generated by local winds. Event-scale upwelling has been demonstrated to have substantial impacts on fisheries [@roy; @van] (RWS: Why is this last sentence here? It has nothing to do with the rest of the paragraph.).

The Cape peninsula is influenced by south-east trade winds during summer and westerlies during winter, thus a pronounced seasonal cycle is evident within patterns of wind movement across the coastline. During spring, summer, and autumn seasons, the anticyclonic high-pressure cell periodically ridges eastwards south of the continent (RWS: Citation). During this process, south easterly winds predominate, with peak frequencies occurring during spring and summer months resulting in more upwelling events during summer and winter months (RWS: Citation). During the winter seasons, low pressure cells are driven from the west to the east as a result of the atmospheric circumpolar waves and so north-westerly winds dominate during winter. South easterly winds dominate during summer but are still present for short durations during winter [@andrews]. Within a range of kilometres off the shore, wind conditions vary due to the effects of mountain funnelling within the region (RWS: Citation).

The dynamics of upwelling within the nearshore (<400m from the coastline) remains poorly understood. Upwelling metrics are often calculated far offshore because measurements are typically taken from large ships that cannot move close to the shore [@andrews]. Cold temperatures are important in describing species population distribution limits and are proven to set limits in the north or southwards higher latitudes (Firth et al., 2011). There has not yet been any understanding of how the upwelling process varies at local versus broader scales. In this study, it was hypothesized that upwelling is more dominant within coastal regions instead of broader scale (RWS: This sentence needs to be re-written for clarity). Very few studies have addressed the changes in the intensity or duration of cold-spells or cold events at varying distances from the coast within the Benguela region (RWS: this should be said before you announce what your hypothesis is).

Strong temperature gradients existing between offshore and inshore oceanic waters may result in different upwelling event readings. Upwelling relatively close to the shore has been found to depend on the strength and direction of the wind and tend to vary based on seasonal wind variation [@andrews]. Inshore temperatures tend to be lower during summer months due to upwelling, but there is an increase in offshore temperatures in the later summer months as a result of warming of the South East Atlantic Surface waters [@andrews; @Shannon]. Due to the temperature gradients that exist inshore and offshore, this study focused on observing upwelling patterns from the coastal region to 50km from the coastline.

The central assumption of this project was that remotely-sensed SST data and *in situ* seawater temperature data may be compared. Remotely sensed SST data and *in situ* seawater temperature data were obtained at different resolutions and so slight changes may be found within the different datasets. The primary aim of the first study was to make use of various remotely sensed SST and *in-situ* collected coastal seawater temperature data. The most appropriate dataset for the measurement of the nearshore temperatures within South Africa was the South African Coastal Temperature Network (SACTN) dataset [@smit]. It was useful to examine whether or not the same upwelling patterns exist at different distances from the coastline and if these patterns that exist are evident within the various datasets. This investigation examined recent trends and variation of upwelling events at different distances from the coastline. (RWS: This is a very thorough introduction section but has potentially too much information. I would recomend that you write out a bullet list for this section with the main and sub-points that you want to make with each paragraph and then see that they all flow from one paragraph to the next. If it can be re-order a bit I think that it would be pretty close to a final version once the needed citations are inserted.)

@@

### Site selection

The South African coastline exhibits a large variation in seawater temperature and is divided into four bioregions [@smit, @mead]. The western region of the coastline is dominated by the Benguela Current forming an Eastern Boundry Upwelling System (EBUS) [@hutchings], which provides a natural laboratory for this study. Annual mean coastal seawater temperatures within this region have a range of 12.3 ± 1.2°C at the Western limit near the Namibian border. Seasonal upwelling is controlled by south-easterly trade winds and as a result reflects distinct temperature variations representing much lower temperatures within the upwelling cells over a fairly narrow continental shelf found from the Cape Peninsula to Cape Columbine. This region represents monthly mean temperatures at generally intermediate at cold and warm temperatures [@luning] (RWS: This sentence is unclear, it would also be good to include a temperature value here, like the previous sentence). In order to accurately examine upwelling patterns at various distances from the coast, several sites from the SACTN dataset along the west coast of South Africa (i.e. sites influenced by the Benguela current) were selected. Since temperature data were not evenly recorded for each of the sites, the entire dataset was narrowed to only those sites with a time series of longer than 30 years. This yielded a total of five sites within the Benguela region (Fig 1 (RWS: The text is too small on this figure. I recommend saving it at a smaller size so that the text appears larger. I would also move the legend over SA rather than having it on the side of the figure.)). (RWS: This sub-section should come after the temperature and wind data sections)

@@

### Temperature data

The South African Coastal Temperature Network (SACTN) data served as the *in situ* temperature dataset used in this study. This dataset consists of coastal seawater temperatures obtained from 129 sites along the South African coastline. The SACTN has collected SST data along the coastline of South Africa from as early as 1972 till present date, with contributions made from several organisations and governmental departments (RWS: The SACTN itself does not collect data. Be careful with your word choice). The temperature data within this dataset were obtained with different methods and instruments, such as hand held thermometers or underwater temperature reorders (UTRs). The precision of the instruments used vary accordingly, with hand-held thermometers recording temperatures at a precision of 0.5°C as opposed to the underwater temperature recorders which have an instrument precision of 0.001°C. The duration and extent of recordings per site were uneven with the longest time series in the dataset being that of Gordons Bay, which started on 13 September 1972 and concluded on 26 January 2017. During the 1970s, a total of 11 time series began recording. A further 53 entries were added during the 1980s, 34 during the 1990s, and 18 entries were added during the 2000s. Recordings are still ongoing at many of these sites.

This study also made use of a range of remotely sensed temperature datasets. Satellites provide a system of large spatial and temporal data retrieval. Remotely sensed SST products are produced by a number of organisations across the globe. (RWS: Consider combining or re-working these first three sentences) The AVHRR-only Optimally-Interpolated Sea Surface Temperature (OISST) was used to determine SST within the study region. The AVHRR datasets have been providing global SSTs for more than four decades (Reynolds and Smith, 1994; Pearce et al 2006). OISST is a global 1/4° gridded daily SST product that assimilates both remotely sensed and *in situ* sources of data to create a level-4 gap free product (Banzon et al., 2016). The second remotely sensed dataset used in this study was the Multi-scale Ultra-high Resolution (MUR) Sea Surface Temperature Analysis. This dataset is produced using satellite instruments from 1 June 2002 to present (PO.DAAC, https://podaac.jpl.nasa.gov/). MUR provides SST data at a global grid of 0.01° in longitude latitude and are currently among the highest resolution SST datasets available. The CMC dataset constitutes the third dataset and is a version 3.0 Group for High Resolution Sea Surface Temperature (GHRSST) Level 4 dataset with a 10km resolution constructed by the Canadian Meteorological Center (CMC). The CMC dataset combines infrared satellite SST at numerous points in the time series from the AVHRR, the European Meteorological Operational-A (METOP-A) and Operational-B (METOP-B) platforms, and microwave SST data from the Advanced Microwave Scanning Radiometer 2 in conjunction with in situ observations of SST from ships and buoys from the ICOADS program (RWS: citation needed).

An advantage of using of *in situ* data over satellite data is that they provide a more accurate representation of the thermal properties closer to the coast where satellite temperatures do not capture temperature properties well and can therefore more accurately explain upwelling within the coastal inshore environment. Smit et al (2013) have shown that satellite data collected along the South African coastline represent a warm bias reaching as high as 6°C over *in situ* temperatures within the nearshore environment. (RWS: Knowing this, why do you say near the end of your intro section that you are okay with assuming that in situ and satellite data are similar along the coast?)

In order to create a time series of each of the remotely sensed SST data and also to compare whether the same upwelling patterns exist on the broad-scale against the *in situ* time series, shore-normal transects were extended from the *in situ* sites along the west coast of South Africa. Temperature values from the various satellite datasets were extracted at 15, 25, and 50 km along the shore-normal transect from the coast  (Fig. 1). The daily SST values obtained from the various remotely sensed datasets along each of the transects were used to produce offshore, broad-scale (RWS: What does broad-scale mean?) time series that could accurately be used as a dataset to be compared against the *in situ* time series.

### Wind data

The wind was an important variable in this study as it was hypothesised that wind direction and wind speed may have a direct influence on upwelling, consequently, the wind was investigated for its impact on upwelling at specific sites along the west coast of South Africa. Wind data were obtained from the South African Weather Service (SAWS), and were provided at a three-hour temporal resolution. The wind variables used here were wind direction (dirw) and wind speed (spw). Since the wind data were modelled at three-hour resolutions, they were converted into daily data points in order to compare them with the temperature data. The `circular()` function in R software (RWS: Cite R) was used to create circular objects around the wave data in order to calculate the daily wind parameters. ERA5 hourly data on pressure levels were used as a second wind source (RWS: Need ERA5 citation and need a sentence explaining what ERA 5 is). This dataset comprised of u and v variables from 1973 to present (RWS: Need to be more specific about what the u and v variables are).

With temperature and wind values now corresponding to their respective sites and dates, the hypothesis testing upwelling patterns at different distances from the coastline could now be tested. To do this, it was important to identify upwelling. (RWS: This needs to be re-worked.)

### Defining and determining upwelling

In order to determine the change in upwelling signals at different distances from the coastline and to test which dataset showed clear, visible representation of these events, it was first necessary to define when upwelling was occurring. To do this it was necessary to determine a threshold value for identifying when upwelling was occurring. Given that upwelling is primarily caused by alongshore, equatorward winds, it was necessary to consider both temperature and wind values. To determine the upwelling condition, an upwelling index was calculated using the formula presented in Fielding and Davis (1989):

@@

In this equation, μ represents the wind speed (m/s) and θ represents the wind direction, which is measured in degrees. The coastal angle is the angle of each of the sites perpendicular to the coastline. The index relies heavily on wind speed and direction data in order to determine the intensity of upwelling likely to occur. Thereafter, the `exceedance()` function within the __`heatwaveR`__ package (RWS: Cite heatwaveR) was used to determine the consecutive number of days the upwelling index was at or above 1. In order to obtain a more approved threshold value, a percentile value was determined (RWS: Was it though? This sentence wouldn't make sense to someone that hadn't also seen the R code). To determine changes within the upwelling values, the `detect_event()` function within the __`heatwaveR`__ package was used. Thereafter, to determine the upwelling event metrics (frequency and intensity) the detect event function was used as a first filter with a threshold at or below the 25th percentile. Values obtained from the upwelling index were used as a second threshold. (RWS: The description of the statistical methodology here is unclear and needs to be re-written) An R script (https://github.com/AmierohAbrahams/Upwelling_MCS) was used to calculate the individual upwelling events for both the *in situ* and remotely sensed SST data. Because upwelling events were calculated relative to percentile exceedances, rather than an absolute definition such as periods with temperatures above a fixed temperature threshold, upwelling events occurred any time of the year with upwelling shown to be more dominant during spring and summer months. For the use of the heatwaveR package it was best to use a time series of 30 years (RWS: No. The 30 years comes from WMO recommendations and is not a requirement for the heatwaveR package specifically).
